Efficiency

### 2018-01-02 DraftExec.tex

\title{Efficiency and acoustic considerations for pumpjets in the context of conventionally powered submarines}
\author{Aidan Morrison}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{figure}
\includegraphics[width=\textwidth]{HelicalLine.png}
\caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated as shown in \parencite{carlton2007}}
\label{fig:HelicalLine.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{PitchDefinition.png}
\caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as shown in \parencite{carlton2007}}
\label{fig:PitchDefinition.png}
\end{figure}

The selection of the exact blade selection that is selected to define pitch must be specified as being at some fraction of the radius from the centre.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \parencite{carlton2007} pages 35-37.

Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.  It is for this reason that a considerable number of vessels actually have variable pitch propellers.  Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.  Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.  This has advantages for ferries which undertake frequent docking manouvres in confined spaces \parencite{MAN2017} \parencite{lewis1988}.

In Figure \ref{fig:ex2} we can see a sample plot of the mtcars data.

\begin{figure}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}

{\centering \includegraphics[width=\maxwidth]{figures/plots-unnamed-chunk-1-1}

}

\end{knitrout}
\caption{An example plot}
\label{fig:ex2}
\end{figure}

\printbibliography

>>>>

\title{Efficiency}
\author{Aidan Morrison}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{figure}
\includegraphics[width=\textwidth]{HelicalLine.png}
\caption{A blade section of a propeller traces a helical line around a cylinder as it is rotated \parencite{carlton2007}}
\label{fig:HelicalLine.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{PitchDefinition.png}
\caption{The definition of pitch is the distance traveled by a blade section along a cylinder, as given in \parencite{carlton2007}}
\label{fig:PitchDefinition.png}
\end{figure}

The selection of the exact blade selection that is selected to define pitch must be specified as being at some fraction of the radius from the centre.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.

Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.  It is for this reason that a considerable number of vessels actually have variable pitch propellers.  Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.  Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.  This has advantages for ferries which undertake frequent docking manouvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).

\begin{figure}
\includegraphics[width=\textwidth]{CPPropeller.png}
\caption{Controllable Pitch Propellers (CP Propellers) allow the pitch of the blades to be varied. Image credit: \parencite{mollard2011}}
\label{fig:CPPropeller.png}
\end{figure}

\subsubsection{Pitch and the Advance Ratio}
Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.

\begin{figure}
\includegraphics[width=\textwidth]{OpenWater.png}
\caption{A typical chart showing open water efficiency of a propeller, in this case a Wageningen B5-75. Image: MARIN}
\label{fig:OpenWater.png}
\end{figure}

This chart also shows two other closely related dimensionless coefficients, the torque and thrust coefficients ($K_Q$ and $K_T$) which we shall not dwell on here.  However, it is important to note that a number of different lines are expressed on this sort of chart, which represent different versions of the same propeller, with only the pitch (expressed in its dimensionless form as $P/D$) being different in each case. The x-axis, $J$, also warrants a clear definition:

\begin{equation}
\label{eq:2}
J = \frac{V_a}{nD}
\end{equation}

Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.  For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related. To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.  Another key distinction for the consideration of jets later on, is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field. In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.

One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$. This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced. This represents circumstances where a propeller encounters effectively zero load.  This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.

A closely related concept here is the slip ratio, which is a measure of the difference between the pitch (the distance the blade section would travel if it traced a perfect helix for one rotation) and the actual distance that the blade did advance through the water.  This is neatly defined and shown in a diagram in \ref{fig:SlipRatio.png}.

\begin{figure}
\includegraphics[width=\textwidth]{SlipRatio.png}
\caption{The slip ratio of a propeller. Image credit: \cite{MAN2017}}
\label{fig:SlipRatio.png}
\end{figure}

It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.  The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water. In this case the slip ratio is one.  This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.  Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.  Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.

\subsection{Ducted Propellers}
A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.  Ducted propellers fall into broadly two categories, one of which is referred to frequently and interchangeably as a pump-jet.  The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating type.  The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).

A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted. (Aside: It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.  But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear.)

\begin{figure}
\includegraphics[width=\textwidth]{AccelDecel.png}
\caption{he two types of ducted propeller, accelerating (left) and decelerating (right) as shown in \parencite[249]{mollard2011}}
\label{fig:AccelDecel.png}
\end{figure}

A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved. In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.  In the decellerating case, the available area contracts after the water has passed the impeller. In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.

\begin{figure}
\includegraphics[width=\textwidth]{DuctStreamlines.png}
\caption{The streamlines of flow through different duct types \parencite[214]{lewis1988}}
\label{fig:DuctStreamlines.png}
\end{figure}

\subsubsection{The Accelerating Duct}
Accelerating ducts are most commonly used by vessels which have to operate at high loadings at very low speed, such as tugs, barges, or trawlers.  In those circumstances, relatively small vessels need to exert considerable forces against substantial loads, at very low water-speeds.  Also, in these circumstances, the practical size of a propeller might be constrained, since operating an optimally sized propeller might increase the complexity of machiner required to deliver the power to so low a hub in the water, and it might also be difficult to arrange for the thrust to be directed substantially against the centre of drag of the vessel.

The effect of the nozzle in these circumstances actually increase the optimal efficiency of the propeller, when operating at very low water-speeds, since the pressure reduction on the inside of the nozzle induced by the movement of the water through the nozzle, actually produces a net thrust forward, since the interior face at the opening of the nozzle is necessarily angled forward. This has the effect of distributing the suction force across a larger area, which now includes the interior of the nozzle.  An equivalent way of describing the same effect is that the nozzle causes the water to be accelerated more gradually, over a longer distance, by inducing it to move more quickly as it passes through to the entrance.  As a broad generalisation, more gradual or gentle actions tend to involve lower overall losses to turbulence, and hence the efficiency of the propeller can be increased.

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation can be diminished.

However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.  In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased, and cavitation can and does still occur in ducted propellers near the blade tips \parencite{moulijn2015}.

\begin{figure}
\includegraphics[width=\textwidth]{CavitatingDuct.png}
\caption{An example of the cavitation occurring at the blade tips of a ducted propeller. Image credit \cite{moulijn2015}}
\label{fig:CavitatingDuct.png}
\end{figure}

A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}. A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.  A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.  At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.  When the exterior of the duct experience relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.  Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).

\begin{figure}
\includegraphics[width=\textwidth]{DuctedEfficiency.png}
\caption{Accelerating ducts produce a net positive thrust at low advance ratios, but a net negative thrust at higher advance ratios as the drag on the nozzle exterior increases. Image credit:Open-water test results of Ka 3-65 screw series with nozzle no. 19A \parencite{oosterveld1970}}
\label{fig:DuctedEfficiency.png}
\end{figure}

\subsubsection{The Decelerating Duct or Pumpjet}
Decelerating ducts necessarily produce a negative thrust, since the process of raising the pressure of the water about the impeller (equivalent to slowing it down) necessitate reducing its kinetic energy and consequently momentum.  In a wide range of literature introducing the fundamentals of ducted propellers and water-jets, or discussing their more modern development there is universal acknowledgement that the choice of a decelerating duct is generally made for the avoidance of cavitation for specialised (generally military) purposes, rather than for the achievement of increased efficiency (\cite{carlton2007}, \cite{oosterveld1970}, \cite{mollard2011}, \cite{haimov2010}, \cite{lewis1988}).

In order to achieve a reduction in cavitation, however, care must be taken in the design of a decelerating duct.  Since the negative thrust produced by the duct must necessitate additional work being done by the impeller, there can be circumstances where the additional loading of the impeller out-weighs the impacts of elevating the pressure around it. \cite[24-25]{oosterveld1970} provides a more thorough derivation of the circumstances in which this can be reliably achieved, which generally include larger blade area ratios or with more rows of rotors for higher loading, or circumstances where the loading of the propeller is quite low.  This creates an inevitable and necessary tension between optimisation of a duct for propulsive efficiency and reduction of cavitation, which will be returned to in subsequent section.

It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.  In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller". Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.  Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.

\begin{figure}
\includegraphics[width=\textwidth]{PumpjetOutline.png}
\caption{In \cite{carlton2007} the pump jet is described as a development of a classical ducted propeller, with the noticeable addition of a stator}
\label{fig:PumpjetOutline.png}
\end{figure}

In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract at this point.

The addition of some new part such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.

\subsubsection{Waterjets}
The use of the word waterjet, as opposed to pumpjet, signifies that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}.

\begin{figure}
\includegraphics[width=\textwidth]{PumpjetWaterjet.png}
\caption{A comparison of a pumpjet and waterjet as given in \cite{wislicenus1973}}
\label{fig:PumpjetWaterjet.png}
\end{figure}

\section{Constraints on the efficiency of pumpjets at low speed}

\subsection{Recent Commentary}
There has been already some attention given to the probable efficiencies of pump jets at low speed. Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.  This argument relies pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.  As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.  This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.

There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017} and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.  Whilst the argument that the state of the art is much advanced is obviously true, it is the case with marine propulsion, as with most sciences, that new developments tend to confirm, rather than annul, many of the workings that have been previously undertaken.  To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.  In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.

\subsection{Modern literature and results}
In addition, more modern works, including those which involve experimental validation of theoretical results also tend to show that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.

A 1995 Japanese experimental study on the performance of a waterjet demonstrated a close experimental fit to a theoretical relationship which sees the propulsive efficiency of the waterjet fall to zero with craft velocity, with significant declines in efficiency commencing just under 10m/s, or somewhere around 18kt \parencite{fujisawa1995}.

\begin{figure}
\includegraphics[width=\textwidth]{EfficiencyJapan.png}
\caption{A graph from a Japanese study of waterjet efficiency shows efficiency rapidly declining below craft velocities of 10m/s, in accordance with theory. \parencite{fujisawa1995}}
\label{fig:EfficiencyJapan.png}
\end{figure}

In 2015 an Australian team including staff from the Australian Maritime College and Incat published results of an experimental study comparing the use of waterjets and propellers for medium-speed ferries operating in the 20-30kt speed range \parencite{kamal2015}.  This study explicitly set out to establish the cross-over point for where propellers become more efficient than jets, since Incat was interested in pioneering highly efficient medium-speed vessels, after having considerable success in very high speed catamarans.  It is again acknowledged as given that at lower speeds propellers are more efficient than waterjets.  The study found that propellers were considerably more efficient over the entire speed range tested, often by a factor of two in terms of transport efficiency.

\begin{figure}
\includegraphics[width=\textwidth]{EfficiencyTransport.png}
\caption{An Australian experimental study aimed at determining the crossover point for the efficiency of jets and propellers in the 20-30kt range revealed much higher power was required across the entire speed range \parencite{kamal2015}}
\label{fig:EfficiencyTransport.png}
\end{figure}

Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.  They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.  However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets). Pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.  Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.

At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role. The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.  These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.  Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.

\begin{figure}
\includegraphics[width=\textwidth]{BMTShape.png}
\caption{The 'fully submerged waterjet' proposed by BMT and Rolls Royce for military surface craft may more closely represent the duct shape used on submarines than waterjets that release above the surface. \parencite{giles2010}}
\label{fig:BMTShape.png}
\end{figure}

The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.  The the vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt. The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form. If the power could be reduced by 5\% (e.g. by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots." \parencite{giles2010}

\begin{figure}
\includegraphics[width=\textwidth]{BMTPowering.png}
\caption{Studies on submerged waterjets for military surface craft currently show that jets require more power than conventional propellers right up to 30kt \parencite{giles2010}}
\label{fig:BMTPowering.png}
\end{figure}

In paper also provides specific curves for the efficiency of the propulsor as a function of vessel speed, which demonstrated that the waterjet efficiency was lower than that of conventional propellers throughout the speed range, but declined particularly precipitously below speeds of 10kt.  At 5kt, the propeller was more than twice as efficient as the waterjet.

\begin{figure}
\includegraphics[width=\textwidth]{BMTEfficiency.png}
\caption{Studies on submerged waterjets for military surface craft currently show that the efficiency of jets decline sharply below 10kt \parencite{giles2010}}
\label{fig:BMTEfficiency.png}
\end{figure}

It appears from recent examples in the literature that there is no evidence of jets approaching the same efficiency as propellers at speeds as low as 10kt.  Indeed, there seems to be considerable evidence that jets of most kinds suffer substantial efficiency penalties relative to propellers in this regime, and that this is understood and accepted as an inevitable design trade-off to be made.  In the case of a submarine, which might spend nearly 100\% of its time travelling at 10kt or less, this disadvantage (as opposed to less than 35\%, as assumed for a surface ship) would become extremely substantial, which will be quantified in later sections.

Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.  This stronly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.  To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.  This should further limit the range of error that might be made in estimating realistic bounds of performance for an unknown system.

\subsection{Theoretical Limitations}
In this section I will appraise some of the available literature, including some of the most canonical works in the design of pumpjets which is now declassified from US Navy funded research (and still extremely frequently cited by contemporary researchers).  Before doing so, I'll attempt to summarise in simple language the mose essential conclusions in simple language.

My broad conclusion is that the optimisation of a system for minimising cavitation, and hence improving acoustic performance, is a separate and independent design requirement to that of optimising a system for propulsive efficiency.  These two design requirements can be found either to be in tension, or alignment, in different regimes.  At high speeds, where cavitation tends to have severe impacts on propulsive efficiency of open propellers, and high mass-flow (which leads to good efficiency) can be achieved through compact ducts and pumps, pumpjets offer very good performance on both metrics.  Hence, in this regime the two design requirements tend to align relatively well in favour of pump-jets.  However, in very low-speed, low-load environments, where cavitation effects are negligible, and larger areas need to be engaged achieve good mass-flow and efficiency, the two design requirements align relatively well in favour of an open propeller.  In a variety of intermediate regimes, the two requirements find themselves in tension.

For this particular consideration, two questions are of paricular importance.  The first is where the clear transitions in the regimes will tend to lie, with respect to the probable requirements of a conventional submarine.  The second is what drives the tension between those design requirements in the intermediate regimes, and whether it is a fundamental or inescapble tension in the laws of physics, or simply a technical hurdle which imminent improvements in the state of the art are likely to overcome by engineering improvements or innovation.

The answer to the first question, as discussed in earlier sections, seems to be that the operating speeds of a submarine tend to lie far below the probable levels where pumpjets are even remotely comparable to waterjets for efficiency.  In later sections I will address further the acoustic comparisons in the speed ranges at 10kt and below.

The answer to the second question seems to be that there the tension between acoustic and propulsive performance at intermediate speeds and loadings is driven by foundational physical requirements.  That is not to say that optimal means of balancing those requirements will not still be found, and improved, as the art advances.  However, it does mean that significantly adapting using known variables that can be changed to favour one design requirement are unlikely to yeild large improvements without some trade-off being made to the other.

The reason for this trade-off occurring can be quite simply attributed to the nature of the shroud, or duct, which surrounds the the rotors of the propulsor, and the effect it has on the flow. If the duct acts to increase the pressure at the rotor, (which reduces cavitation) it necessarily must slow the flow at/around the rotor, which necessitates produceing a negative thrust on the duct.  The system as a whole produces a positive thrust, becuase the impeller or rotor pushes a larger positive thrust, but the duct necessarily produces a negative thrust.  At the speeds get lower, and the thrust required gets much much lower, this negative thrust becomes proportionally more significant, as described later.

Put another way, the work required to slow the water down, then speed it up again, becomes a gradually greater penalty the the slower the water is coming in, and the less you actually want to speed it up.  If the water is slowed enough by the duct to effectively stop, then the impeller must do the work of speeding it up (sucking it into the duct) before the duct slows it down, before the impeller raises its pressure, and the pressure accelerates it out of the nozzle afterwards. Compared to simply accelerating the water in one simple push, as a propeller blade does in an unconstrained flow, the efficiency penalty is very large.

\subsection{The impact of duct loss on an ideal propeller}

Derivations of the 'ideal' efficiencies are frequently made for propellers, and jets using basic conservation laws (momentum theory) which consider simply acceleration of some column of water as actuated by a disc representing the impeller or propeller.  A fuller description of these equations can be found in  most standard texts \parencite[131]{lewis1988}. These methods are computationally simple, and allow simple upper-bounds to be be set for efficiencies achievable for given designs under certain conditions, without considering the portential impact of other more complex effects such as skin friction over the blades, potential flow separation, or cavitation.  As such, they can be easily relied upon to constrain the overall plausible range of values that can be achieved from universally applicable inputs, such as the velocity of water going into a duct, and the velocity of water exiting a nozzle.

Such methods are used in \cite[8-14]{wislicenus1973} to construct an 'ideal' jet efficiency that applies to waterjets and jetpumps. Whilst I won't repeat the full expansion here, but a fundamentally important early conclusion of the derivation is the distinction between 'useful work' and the total energy expended. since the rate of change in momentum is known to be necessarily equal to the force generated, and increasing momentum change (hence thrust) can be achieved either accelerating more mass, or accelerating mass to higher speeds.   Given that energy embodied in a flow is proportional to it's velocity squared, but its momentum only to its velocity, a necessary and inevitable consequence is that some energy is always wasted in accelerating a fluid to a higher velocity than its surrounds.

This is expressed in the equation which represents 'ideal jet efficiency', which is directly equivalent to 'ideal propeller efficiency' using momentum theory of propeller action:

\begin{equation}
\label{eq:3}
\eta_j = \frac{1}{1+\frac{\Delta V}{2V_0}}
\end{equation}

The inevitable consequence of this relationshiop is that efficency approaches unity when $\Delta V$ (the change in water of the water between entering and exiting the system) relative to $V_0$ approaches zero.  Intuitive embodiments of this relationship are (equivalently) that efficiency is maximum when the least force is exerted on any given bit of water, which might be when negligibly small work is being done (extreme low speed) or when such a large mass of water is acted on that it barely needs to be pushed at all.  This relationship would lead to the conclusion that infinitely large propellers or jets are always desirable from an efficiency perspective.  This of course isn't practical to build, and also doesn't hold when the effects of the drag on the propeller blades, or duct, are taken into account.  Infinitely large systems would increase the total weight, as well as having significant surface draw when moving through the water, all of which are neglected in this 'ideal' consideration.

In \cite[8-14]{wislicenus1973}, Wislicenus expands the relationship in the case of the jet to incorporate terms to represent some of the inevitable adjustments that would occur in the case of pump jets and waterjets, particulalry the duct head loss ($K$) (fraction of energy required to push the water through the duct), as well as the additional drag on the outside of the propulsor ($\Delta T$).  This leads to an adjusted efficiency relationship, which still neglects any losses due to pump efficiency, or change in height of the jet (which is added for consideration for waterjets, but not relevant here).

\begin{equation}
\label{eq:4}
\eta_j = \frac{1}{1 + \frac{\Delta V}{2V_0} + K\frac{V_0}{2 \Delta V}}(1 - \frac{\Delta T}{T})
\end{equation}

The conclusion of the analysis, and the plotting of these results for certain possible values of $K$ and $\Delta T$ is: "Its outstanding characteristic is that this efficiency approaches zero rather than unity for $\Delta V / V_0 = 0$, even for small duct losses".

\begin{figure}
\includegraphics[width=\textwidth]{PumpjetEfficiency.png}
\caption{Ideal jet efficiencies sharply appoach zero at low values of $\Delta V / V_0$, where as ideal propeller efficiencies without a duct $K = 0$ approach unity. Credit: \cite{wislicenus1973}}
\label{fig:PumpjetEfficiency.png}
\end{figure}

Wislicenus also acknowledges \cite{wislicenus1973} another author, Brandau, who makes a different assumption about the relationship between the duct loss and water velocity \parencite{brandau1967} in his derivation (along with another thorough discussion of potential and ideal efficiencies), who produces a different peak efficiency point, but retains a similar overall shape featuring efficiency falling very rapidly to zero at low values of $\Delta V / V_0$, and contrasting this with the ideal propeller, which approaches unity in this low regime.

It's important to note at this point that whilst charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.  For the avoidance of any confusion, for a particular fixed vessel size and drag, and fixed propulsion system design smaller values of $\Delta V / V_0$ will correspond to to the propulsion system exerting less thrust, and also moving at a lower speed under self-propulsion, since the thust required (proportional to $\Delta V$ when operating at steady state) will decline faster than the vessel speed or water speed at intake, since the drag on the vessel scales with the square of the vessel speed.

\begin{figure}
\includegraphics[width=\textwidth]{EfficiencyBrandau.png}
\caption{Alternative derivations of ideal jet efficiency also show sharply declining efficiency for low values of $\Delta V / V_0$ Credit: \cite{brandau1967}}
\label{fig:EfficiencyBrandau.png}
\end{figure}

The structure of these overall curves is found to be overall quite similar for both pumpjets and waterjets, though the probably values for the losses incurred in the intake duct, or on the exterior of the shroud, are obviously different \parencite[13-14]{wislicenus1973}.  Similar efficiency shapes can be found for discussions focussing explicitly on underwater pumpjets \parencite[13]{henderson1964}. It should be noted that the theoretical efficiency derived by these general equations is consistent both in derivation and results with those produced by other authors \parencite[227]{lewis1988}.

Care should be taken, however, to note that different graphs for efficiency can be produced which have different variables on the x axis.  For example, \cite[247]{mollard2011} a graph of ideal efficiency of a pump jet is shown using the speed of advance divided by the jet speed, or $V_1/V_2$.  Given that the x axis used in the previous charts used $\Delta V / V_0$, which would be related to $V_2-V_1$ in this case, this has the effect of reversing the effective direction of the chart. In this case, for a fixed jet and vessel design, high-speed operations will be towards the left of the chart, and the precipitous decline in efficiency will occur as speed declines to zero, along with thrust, toward the right of the chart, as opposed to the left in earlier cases.

\begin{figure}
\includegraphics[width=\textwidth]{EfficiencyMollard.png}
\caption{Other modern reference texts might demonstrate jet or propeller efficiencies with different x-axes, effectively reversing the way the graph should be understood for a single system operating at different speeds. Credit:  \cite{mollard2011}}
\label{fig:EfficiencyMollard.png}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{EfficiencyBulten.png}
\caption{Another graph representing efficiency as a function of the ratio of jet velocity to incoming velocity, which shows efficiency declining at low thrust (to the right).  The impact of increasing duct losses can be clearly seen, as opposed to the 'ideal' propeller without a duct. Credit: \cite{bulten2006numerical}}
\label{fig:EfficiencyBulten.png}
\end{figure}

Similarly, it is very important to note that the very frequently used x-axis unit $J = V_a/nD$ when showing efficiency curves for both propellers and ducted propellers (such as used in earlier figures) are of a similar orientation to the chart shown in the figure above.  The left side of the chart will correspond to the highest number of rotations for a given amount of progress through the water.  For vessel of fixed load, and a propeller of fixed geomtry, this will correspond to a propeller facing the higest drag force due to the vessel moving at the highest speed.
